```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(readr)          # Data Input
library(tidymodels)     # Data Manipulation
library(lubridate)      # Data Manupulation
library(dplyr)          # Data Manipulation
library(reshape2)       # Data Manipulation
library(caTools)        # Data Manipulation
library(corrplot)       # Data Visualisation
library(ggplot2)        # Data Visualization
library(viridis)        # Data Visualization
library(ggthemes)       # Data Visualization
library(pROC)           # Metrics
library(caret)          # Machine Learning
library(xgboost)        # xgboost model
```

This practical is based on exploratory data analysis and prediction of a dataset derived from a municipal database of healthcare administrative data. This dataset is derived from Vitoria, the capital city of Espírito Santo, Brazil (population 1.8 million) and was freely shared under a creative commons license.

**Generate an rmarkdown report that contains all the necessary code to document and perform: EDA, prediction of no-shows using XGBoost, and an analysis of variable/feature importance using this data set. Ensure your report includes answers to any questions marked in bold. Please submit your report via brightspace as a link to a git repository containing the rmarkdown and compiled/knitted html version of the notebook.**

## Introduction

The Brazilian public health system, known as SUS for Unified Health System in its acronym in Portuguese, is one of the largest health system in the world, representing government investment of more than 9% of GDP. However, its operation is not homogeneous and there are distinct perceptions of quality from citizens in different regions of the country.  Non-attendance of medical appointments contributes a significant additional burden on limited medical resources.  This analysis will try and investigate possible factors behind non-attendance using an administrative database of appointment data from Vitoria, Espírito Santo, Brazil.

The data required is available via the [course website](https://github.com/maguire-lab/health_data_science_research/tree/master/static_files/practicals/lab1_data).

### Understanding the data

**1** Use the data dictionary describe each of the variables/features in the CSV in your report.
PatientID: Unique identifier for each patient.
Answer-
AppointmentID: Unique identifier to each appointment.
Gender: Patient Gender (limited to Male or Female).
ScheduledDate: date of the actual appointment.
AppointmentDate: date at which the appointment supposed to take place.
Age: Patient's age.
Neighbourhood: District of Vitória in which the appointment.
SocialWelfare: Patient is a recipient of Bolsa Família welfare payments.
Hypertension: Patient previously diagnosed with hypertension (Boolean).
Diabetes: Patient previously diagnosed with diabetes (Boolean).
AlcoholUseDisorder: Patient previously diagnosed with alcohol use disorder (Boolean).
Disability: Patient previously diagnosed with a disability (severity rated 0-4).
SMSReceived: At least 1 reminder text sent before appointment (Boolean).
NoShow: Patient did not attend scheduled appointment (Boolean: Yes/No).


**2** Can you think of 3 hypotheses for why someone may be more likely to miss a medical appointment?
Answer-
Hypotheses 1: Patients living in remote neighborhoods may face transportation challenges or longer travel times, leading to higher rates of non-attendance.
Hypotheses 2: Patients with chronic conditions such as hypertension, diabetes, or disabilities might have more frequent appointments and might miss some due to health complications.
Hypotheses 3: Patients who do not receive SMS reminders might forget their appointments or be less motivated to attend.

**3** Can you provide 3 examples of important contextual information that is missing in this data dictionary and dataset that could impact your analyses e.g., what type of medical appointment does each `AppointmentID` refer to?
Answer-
Example 1- Information about whether the appointment was for a general check-up, specialist consultation, or follow-up visit. This can impact the perceived importance and urgency of the appointment.
Example 2- Information on how far patients have to travel to reach the healthcare facility. This could provide insight into whether travel distance impacts non-attendance.
Example 3- Detailed information about the socioeconomic status of the patients such as income level, employment status, and education level, which could influence their ability to attend appointments.

## Data Parsing and Cleaning

**4** Modify the following to make it reproducible i.e., downloads the data file directly from version control

```{r parse}
url <- 'https://raw.githubusercontent.com/maguire-lab/health_data_science_research_2024/main/lab1_data/2016_05v2_VitoriaAppointmentData.csv'

raw.data <- read_csv("2016_05v2_VitoriaAppointmentData.csv", col_types = 'fffTTifllllflf')

head(raw.data)
```

Now we need to check data is valid: because we specified col_types and the data parsed without error most of our data seems to at least be formatted as we expect i.e., ages are integers

```{r}
raw.data %>% filter(Age > 110)
```
We can see there are 2 patient's older than 110 which seems suspicious but we can't actually say if this is impossible.

**5** Are there any individuals with impossible ages? If so we can drop this row using `filter` i.e., `data <- data %>% filter(CRITERIA)`
Answer- There are individuals with SocialWelfare and SMSReceived who are too young.
```{r}
raw.data %>% filter(Age < 2) %>% filter(SocialWelfare == TRUE)
```

## Exploratory Data Analysis
First, we should get an idea if the data meets our expectations, there are newborns in the data (`Age==0`) and we wouldn't expect any of these to be diagnosed with Diabetes, Alcohol Use Disorder, and Hypertension (although in theory it could be possible).  We can easily check this:

```{r}
raw.data %>% filter(Age == 0) %>% select(Hypertension, Diabetes, AlcoholUseDisorder) %>% unique()
```

We can also explore things like how many different neighborhoods are there and how many appoints are from each? 

```{r}
count(raw.data, Neighbourhood, sort = TRUE)
```
**6** What is the maximum number of appointments from the same patient?
```{r}
count(raw.data, PatientID, sort = TRUE)
```
Answer- The maximum number of appointments from a patient is 88 with ID 822145925426128.

Let's explore the correlation between variables:

```{r}

# let's define a plotting function
corplot = function(df){
  
  cor_matrix_raw <- round(cor(df),2)
  cor_matrix <- melt(cor_matrix_raw)
  
  
  #Get triangle of the correlation matrix
  #Lower Triangle
  get_lower_tri<-function(cor_matrix_raw){
    cor_matrix_raw[upper.tri(cor_matrix_raw)] <- NA
    return(cor_matrix_raw)
  }
  
  # Upper Triangle
  get_upper_tri <- function(cor_matrix_raw){
    cor_matrix_raw[lower.tri(cor_matrix_raw)]<- NA
    return(cor_matrix_raw)
  }
  
  upper_tri <- get_upper_tri(cor_matrix_raw)
  
  # Melt the correlation matrix
  cor_matrix <- melt(upper_tri, na.rm = TRUE)
  
  # Heatmap Plot
  cor_graph <- ggplot(data = cor_matrix, aes(Var2, Var1, fill = value))+
    geom_tile(color = "white")+
    scale_fill_gradient2(low = "darkorchid", high = "orangered", mid = "grey50", 
                         midpoint = 0, limit = c(-1,1), space = "Lab", 
                         name="Pearson\nCorrelation") +
    theme_minimal()+ 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                     size = 8, hjust = 1))+
    coord_fixed()+ geom_text(aes(Var2, Var1, label = value), color = "black", size = 2) +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank(),
      axis.ticks = element_blank())+
      ggtitle("Correlation Heatmap")+
      theme(plot.title = element_text(hjust = 0.5))
  
  cor_graph
}

numeric.data = mutate_all(raw.data, function(x) as.numeric(x))

# Plot Correlation Heatmap
corplot(numeric.data)

```

Correlation heatmaps are useful for identifying linear relationships between variables/features.
In this case, we are particularly interested in relationships between `NoShow` and any specific variables.

**7** Which parameters most strongly correlate with missing appointments (`NoShow`)?
Answer- SMSReceived and the appointment ID correlate most strongly.

**8** Are there any other variables which strongly correlate with one another?
Answer- Yes there are and these are SMSReceived, Diabetes and Hypertension.

**9** Do you see any issues with PatientID/AppointmentID being included in this plot? 
Answer- 'PatientID' and 'AppointmentID' in the correlation plot can be an issue because these are unique identifiers and not continuous variables.

Let's look at some individual variables and their relationship with `NoShow`.

```{r,fig.align="center"}
ggplot(raw.data) + 
  geom_density(aes(x=Age, fill=NoShow), alpha=0.8) + 
  ggtitle("Density of Age by Attendence")
```
There does seem to be a difference in the distribution of ages of people that miss and don't miss appointments.  
However, the shape of this distribution means the actual correlation is near 0 in the heatmap above. This highlights the need to look at individual variables.

Let's take a closer look at age by breaking it into categories.

```{r, fig.align="center"}
raw.data <- raw.data %>% mutate(Age.Range=cut_interval(Age, length=10))

ggplot(raw.data) + 
  geom_bar(aes(x=Age.Range, fill=NoShow)) + 
  ggtitle("Amount of No Show across Age Ranges")

ggplot(raw.data) + 
  geom_bar(aes(x=Age.Range, fill=NoShow), position='fill') + 
  ggtitle("Proportion of No Show across Age Ranges")

```

**10** How could you be misled if you only plotted 1 of these 2 plots of attendance by age group?
Answer- In plot 1, there are significant differences in the total number of appointments made across different age groups, which can skew the count of missed appointments. In plot 2, there are small sample sizes within certain age ranges, causing a misunderstanding that may suggest significant differences in missed appointment rates when differences are not statistically significant.

The key takeaway from this is that  number of individuals > 90 are very few from plot 1 so probably are very small so unlikely to make much of an impact on the overall distributions. 
However, other patterns do emerge such as 10-20 age group is nearly twice as likely to miss appointments as the 60-70 years old.

Next, we'll have a look at `SMSReceived` variable:

```{r,fig.align="center"}
ggplot(raw.data) + 
  geom_bar(aes(x=SMSReceived, fill=NoShow), alpha=0.8) + 
  ggtitle("Attendance by SMS Received")

ggplot(raw.data) + 
  geom_bar(aes(x=SMSReceived, fill=NoShow), position='fill', alpha=0.8) + 
  ggtitle("Proportion Attendance by SMS Received")
```


**11** From this plot does it look like SMS reminders increase or decrease the chance of someone not attending an appointment? Why might the opposite actually be true (hint: think about biases)? 
Answer- Seeing the plots, it seems that patients who received SMS did not show up in much higher proportion compared to the opposite.
Now, there are reasons why the opposite might be true. Patients who receive SMS reminders might be more engaged with their healthcare and more likely to attend appointments regardless of the reminder. Thus, the SMS reminders are sent to those who are already less likely to attend, which may not significantly impact their attendance behavior.

**12** Create a similar plot which compares the the density of `NoShow` across the values of disability 

```{r}
ggplot(raw.data) +
  geom_density(aes(x = NoShow, fill = Disability), alpha = 0.8) +
  ggtitle("Density of NoShow by Disability") 
```

Now let's look at the neighbourhood data as location can correlate highly with many social determinants of health. 

```{r, fig.align="center"}
ggplot(raw.data) + 
  geom_bar(aes(x=Neighbourhood, fill=NoShow)) + 
  theme(axis.text.x = element_text(angle=45, hjust=1, size=5)) + 
  ggtitle('Attendance by Neighbourhood')


ggplot(raw.data) + 
  geom_bar(aes(x=Neighbourhood, fill=NoShow), position='fill') + 
  theme(axis.text.x = element_text(angle=45, hjust=1, size=5)) + 
  ggtitle('Proportional Attendance by Neighbourhood')
```

Most neighborhoods have similar proportions of no-show but some have much higher and lower rates.

**13** Suggest a reason for differences in attendance rates across neighbourhoods.
Answer-
One reason can be transportation and accessibility. Neighbourhoods with limited access to transportation or healthcare facilities may have higher no-show rates due to difficulties in traveling to appointments.

Now let's explore the relationship between gender and NoShow.
```{r, fig.align="center"}
ggplot(raw.data) + 
  geom_bar(aes(x=Gender, fill=NoShow))+
  ggtitle("Gender by attendance")

ggplot(raw.data) + 
  geom_bar(aes(x=Gender, fill=NoShow), position='fill')+
  ggtitle("Proportion Gender by attendance")

```

**14** Create a similar plot using `SocialWelfare`

```{r ,fig.align="center"}
ggplot(raw.data) + 
  geom_bar(aes(x=SocialWelfare, fill=NoShow))
  ggtitle("Social Welfare Usage by attendance")

ggplot(raw.data) +
  geom_bar(aes(x=SocialWelfare, fill=NoShow), position='fill') + 
  ggtitle("Proportional Social Welfare Usage by Attendance")
```

Far more exploration could still be done, including dimensionality reduction approaches but although we have found some patterns there is no major/striking patterns on the data as it currently stands.

However, maybe we can generate some new features/variables that more strongly relate to the `NoShow`.

## Feature Engineering

Let's begin by seeing if appointments on any day of the week has more no-show's. Fortunately, the `lubridate` library makes this quite easy!

```{r}
raw.data <- raw.data %>% mutate(AppointmentDay = wday(AppointmentDate, label=TRUE, abbr=TRUE), 
                                 ScheduledDay = wday(ScheduledDate,  label=TRUE, abbr=TRUE))

ggplot(raw.data) +
  geom_bar(aes(x=AppointmentDay, fill=NoShow)) +
  ggtitle("Amount of No Show across Appointment Day") 

ggplot(raw.data) +
  geom_bar(aes(x=AppointmentDay, fill=NoShow), position = 'fill') +
  ggtitle("Proportion of No Show across Appointment Day") 

```
Let's begin by creating a variable called `Lag`, which is the difference between when an appointment was scheduled and the actual appointment.

```{r, fig.align="center"}
raw.data <- raw.data %>% mutate(Lag.days=difftime(AppointmentDate, ScheduledDate, units = "days"),
                                Lag.hours=difftime(AppointmentDate, ScheduledDate, units = "hours"))

ggplot(raw.data) + 
  geom_density(aes(x=Lag.days, fill=NoShow), alpha=0.7)+
  ggtitle("Density of Lag (days) by attendance")
```

**15** Have a look at the values in lag variable, does anything seem odd?
Answer- 
There are extreme values which indicate that many patients did not show up. The negative lag indicates the waiting for hours.

## Predictive Modeling

Let's see how well we can predict NoShow from the data. 

We'll start by preparing the data, followed by splitting it into testing and training set, modeling and finally, evaluating our results. For now we will subsample but please run on full dataset for final execution.


```{r}
### REMOVE SUBSAMPLING FOR FINAL MODEL
data.prep <- raw.data %>% select(-AppointmentID, -PatientID) #%>% sample_n(10000)

set.seed(42)
data.split <- initial_split(data.prep, prop = 0.7)
train  <- training(data.split)
test <- testing(data.split)
```

Let's now set the cross validation parameters, and add classProbs so we can use AUC as a metric for xgboost.

```{r}
fit.control <- trainControl(method="cv",number=3,
                           classProbs = TRUE, summaryFunction = twoClassSummary)
```

**16** Based on the EDA, how well do you think this is going to work?
Answer- 
There is an imbalance in the NoShow variable with more attended appointments than missed appointments.This imbalance may impact the predictive model's performance. The variables that showed correlations are likely to be important predictors for the model. Anomalies, such as extreme values in Lag.days were seen. Addressing these issues and ensuring clean data preprocessing are essential for model performance. Therefore, careful feature selection will be be necessary to achieve accurate predictions.

Now we can train our XGBoost model
```{r}
xgb.grid <- expand.grid(eta=c(0.05),
                       max_depth=c(4),colsample_bytree=1,
                       subsample=1, nrounds=500, gamma=0, min_child_weight=5)

xgb.model <- train(NoShow ~ .,data=train, method="xgbTree",metric="ROC",
                  tuneGrid=xgb.grid, trControl=fit.control)

xgb.pred <- predict(xgb.model, newdata=test)
xgb.probs <- predict(xgb.model, newdata=test, type="prob")
```

```{r}
test <- test %>% mutate(NoShow.numerical = ifelse(NoShow=="Yes",1,0))
confusionMatrix(xgb.pred, test$NoShow, positive="Yes")
paste("XGBoost Area under ROC Curve: ", round(auc(test$NoShow.numerical, xgb.probs[,2]),3), sep="")
```

This isn't an unreasonable performance, but let's look a bit more carefully at the correct and incorrect predictions,

```{r ,fig.align="center"}
xgb.probs$Actual = test$NoShow.numerical
xgb.probs$ActualClass = test$NoShow
xgb.probs$PredictedClass = xgb.pred
xgb.probs$Match = ifelse(xgb.probs$ActualClass == xgb.probs$PredictedClass,
                         "Correct","Incorrect")
# [4.8] Plot Accuracy
xgb.probs$Match = factor(xgb.probs$Match,levels=c("Incorrect","Correct"))
ggplot(xgb.probs,aes(x=Yes,y=Actual,color=Match))+
  geom_jitter(alpha=0.2,size=0.25)+
  scale_color_manual(values=c("grey40","orangered"))+
  ggtitle("Visualizing Model Performance", "(Dust Plot)")
```


Finally, let's close it off with the variable importance of our model:

```{r,fig.align="center"}
results = data.frame(Feature = rownames(varImp(xgb.model)$importance)[1:10],
                     Importance = varImp(xgb.model)$importance[1:10,])

results$Feature = factor(results$Feature,levels=results$Feature)


# [4.10] Plot Variable Importance
ggplot(results, aes(x=Feature, y=Importance,fill=Importance))+
  geom_bar(stat="identity")+
  scale_fill_gradient(low="grey20",high="orangered")+
  ggtitle("XGBoost Variable Importance")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**17** Using the [caret package](https://topepo.github.io/caret/) fit and evaluate 1 other ML model on this data.
```{r}
set.seed(107)
inTrain <- createDataPartition(y = data.prep$NoShow, p = 0.85, list = FALSE)

training <- data.prep[inTrain,]
testing <- data.prep[-inTrain,]

plsFit <- train (
  NoShow ~ .,
  data = training,
  method = "pls",
  preProc = c("center", "scale")
)

plsClasses <- predict(plsFit, newdata = testing)
plsProbs <- predict(plsFit, newdata = testing, type = "prob")

head(plsProbs)
```

**18** Based on everything, do you think we can trust analyses based on this dataset? Explain your reasoning.
Answer-
Based on everything, I do not think we can trust our analyses on this data. The data has a number of issues and that is the extreme values related to age, and not mentioning the type of appointments made.

## Credits

This notebook was based on a combination of other notebooks e.g., [1](https://www.kaggle.com/code/tsilveira/applying-heatmaps-for-categorical-data-analysis), [2](https://www.kaggle.com/code/samratp/predict-show-noshow-eda-visualization-model), [3](https://www.kaggle.com/code/andrewmvd/exploring-and-predicting-no-shows-with-xgboost/report)